Gravitational wave sources in our Galactic backyard: Predictions for BH and NS binaries in LISA

### 2021-06-30 paper/tex/sections/methods.tex

Each DCO loses orbital energy to gravitational waves throughout its lifetime. This causes the binary to shrink and circularise over time. In order to assess the detectability of a binary, we need to know its eccentricity and frequency at the time of the LISA mission. For each binary in our simulated Milky Way, we know that the time from DCO formation to today is $\tau - t_{\rm evolve}$ and that the initial eccentricity and semi-major axis are $e_{\rm DCO}$ and $a_{\rm DCO}$. We find the eccentricity of the binary at the start of the LISA mission, $e_{\rm LISA}$, by numerically integrating its time derivative \citep[][Eq. 5.13]{Peters+1964} given the initial conditions. This can be converted to the semi-major axis at the start of LISA, $a_{\rm LISA} $\citep[][Eq. 5.11]{Peters+1964}, which in turn gives the orbital frequency, $f_{\rm orb, LISA}$, by Kepler's third law.

@@

\subsubsection{Detection rate calculation}
For each physics variation model and DCO type, we compute the total number of merging DCOs in the Milky Way today, $N_{\rm MW}$, based on the COMPAS simulation. This takes into account the true mass and IMF of the Milky Way since our simulations may each have a slightly different galaxy mass and only ever sample from massive stars rather than the full IMF. For a more in-depth discussion of this process see Appendix~\ref{app:rate_normalisation}.

>>>>

Each binary loses orbital energy to gravitational waves throughout its lifetime. This causes the binary to shrink and circularise over time. In order to assess the detectability of a binary, we need to know its eccentricity and frequency at the time of the LISA mission. For each binary in our simulated Milky Way, we know that the time from DCO formation to today is $\tau - t_{\rm evolve}$ and that the initial eccentricity and semi-major axis are $e_{\rm DCO}$ and $a_{\rm DCO}$. We find the eccentricity of the binary at the start of the LISA mission, $e_{\rm LISA}$, by numerically integrating its time derivative \citep[][Eq. 5.13]{Peters+1964} given the initial conditions. This can be converted to the semi-major axis at the start of LISA, $a_{\rm LISA} $\citep[][Eq. 5.11]{Peters+1964}, which in turn gives the orbital frequency, $f_{\rm orb, LISA}$, by Kepler's third law.

@@

\subsubsection{Detection rate calculation}
For each physics variation model and DCO type, we compute the total number of merging DCOs in the Milky Way today, $N_{\rm MW}$, based on the COMPAS simulation. This takes into account the true mass and IMF of the Milky Way since our simulations may each have a slightly different galaxy mass and only ever sample from massive stars rather than the full IMF. For more details see Appendix~\ref{app:rate_normalisation}.

### 2021-06-30 paper/tex/sections/results.tex
update detection rate plot

In this section we present our main results for the detectable LISA DCO population. We find that on average, for our fiducial model, a 4-year LISA mission will detect \BHBHFourYear{} BHBHs, \BHNSFourYear{} BHNSs and \NSNSFourYear{} NSNSs. We first show the distribution of the sources on the sensitivity curve in Section~\ref{sec:dcos_on_sc}, before exploring the variations in the detection rate over different physics variations in Section~\ref{sec:detection_rate_analysis} and analysing the parameter distributions for detectable sources in the fiducial model in Section~\ref{sec:fiducial_distributions}.

@@

\begin{figure*}[tp]
    \centering
    \includegraphics[width=\textwidth]{2_dcos_on_sc.png}
    \caption{Density distribution of detectable DCOs plotted over the LISA sensitivity curve, where each panel corresponds to \textbf{top:} combined density plot for all DCO types (BHBH, BHNS and NSNS) \textbf{bottom:} three panels with individual density distribution of different DCO types. In each panel, we plot the total signal of binaries at their dominant frequency $n f_{\rm orb}$, such that $n$ is the harmonic that produces the most relative gravitational wave luminosity ($n = 2$ for circular binaries). If the density of points is below our lowest contour (2\%) then we plot the points as scatter points, where their sizes corresponds to their STROOPWAFEL weights. The inset colourbars indicate the percentage of the population represented by each contour and the annotated mass is the average chirp mass for all binaries in the panel, which is used in plotting the grid lines. We plot diagonal lines of constant distance, where the straight lines show the signal for a circular binary of average chirp mass, whilst the curved line shows the signal for an eccentric binary with $e = 0.98$. The vertical lines indicate the inspiral time for a circular binary with the average chirp mass. In the top panel we overlay the LISA verification binaries from \citet{Kupfer+2018}. In the bottom panels we add the 90\% contour line for only the circular sources to show how the distribution changes without eccentric sources present.}
    \label{fig:dcos_on_sc}
\end{figure*}

We illustrate the distribution of detectable DCOs on the LISA sensitivity curve in Figure~\ref{fig:dcos_on_sc}. This shows that the detectable population of these massive DCOs is concentrated at comparatively lower frequencies than the LISA verification binaries (shown as stars in the top panel) and the more numerous WDWD population \citep[e.g.][]{Korol+2017}. This is expected since producing the same SNR as a BHBH, BHNS or NSNS with a relatively lower mass (circular) WDWD requires a higher frequency. This finding is in agreement with \citet{Sesana+2020} and as noted in that work, this could possibly be used to distinguish these more massive DCOs from WDWDs. There are several other notable features in these distributions and we plot grid lines of constant distance and inspiral time to help to explain the shape of the distribution.

The straight diagonal lines show where a binary at a fixed distance and with the average chirp mass (annotated in each panel) would lie on the sensitivity curve for different frequencies. We would therefore expect that if a population was entirely circular, it should be bounded approximately between the $0.1$--$30 \unit{kpc}$ lines (roughly the minimum and maximum distance to a source in the Milky Way). From inspection of the bottom panels with each individual DCO type we see that, though this is the true for a large fraction of the population, there is a distinct subpopulation of binaries that extend downwards, especially around $2 \unit{mHz}$. This offshoot is composed of eccentric binaries for which the circular distance contours do not apply. For instance, we plot the 90\% contour of only the circular sources in our sample over the density distribution in each of the bottom panels and it is clear that the circular subsample is bounded by the distance lines. We also plot a line of constant distance at $30 \unit{kpc}$ for an eccentric binary with $e = 0.98$ to show the differences for eccentric sources. Additionally, we note that the peak of the density distribution coincides with the centre of the Milky Way as expected, since binaries are most likely to be formed towards the centre of the Galaxy.

@@

\begin{figure*}[p]
    \centering
    \includegraphics[width=\textwidth]{3_dco_detections.pdf}
    \caption{The number of expected detections in the LISA mission for different DCO types and model variations. Error bars show the 50\% (solid) and 90\% (dotted) confidence intervals. The left axis and grid lines show the number of detections in a 4-year LISA mission and the right axis shows an approximation of the number of detections in a 10-year mission (we scale the axis by $\sqrt{T_{\rm obs}}$, see Table~\ref{tab:detection_rates} for exact rates). Each model is described in further detail in Table~\ref{tab:physics_variations} and details of the fiducial assumptions are in Section~\ref{sec:fiducial_physics}. \todo{subject to change with the updated/new models and new uncertainty estimates}}
    \label{fig:detection_rates}
\end{figure*}

>>>>

In this section we present our main results for the detectable LISA DCO population. We find that on average, for our fiducial model, a 4-year LISA mission will detect about \BHBHFourYear{} BHBHs, \BHNSFourYear{} BHNSs and \NSNSFourYear{} NSNSs (c.f.\ Table~\ref{tab:detection_rates}). We first show the distribution of the sources together with the sensitivity curve in Section~\ref{sec:dcos_on_sc}, before exploring the variations in the detection rate over different physics variations in Section~\ref{sec:detection_rate_analysis} and analysing the parameter distributions for detectable sources in the fiducial model in Section~\ref{sec:fiducial_distributions}.

@@

\begin{figure*}[tp]
    \centering
    \includegraphics[width=\textwidth]{2_dcos_on_sc.png}
    \caption{Density distribution of detectable BHBH, BHNS and NSNS binaries are shown together with the LISA sensitivity curve. In the top panel we show all systems with the LISA verification binaries over plotted (star symbols, \citealp{Kupfer+2018}). In the bottom panels we separate by type. Contours show the percentage of the population enclosed. The remaining 2\% of the population is shown as dots with a size that scales with the sampling weight. For reference we show lines where a circular binary of average mass $\avg{\mathcal{M}_c}$ would reside for a given remaining inspiral time (vertical lines) and distance (diagonal line). To highlight the role of eccentricity we further show the signal expected for an eccentric binary at 30 kpc. The coloured line in the bottom panels shows a contour that encloses 90\% of the population that is circular. See Sec.~\ref{sec:dcos_on_sc} for a discussion.}
    \caption{Density distribution of detectable DCOs plotted over the LISA sensitivity curve, where each panel corresponds to \textbf{top:} combined density plot for all DCO types (BHBH, BHNS and NSNS) \textbf{bottom:} three panels with individual density distribution of different DCO types. In each panel, we plot the total signal of binaries at their dominant frequency $n f_{\rm orb}$, such that $n$ is the harmonic that produces the most relative gravitational wave luminosity ($n = 2$ for circular binaries). If the density of points is below our lowest contour (2\%) then we plot the points as scatter points, where their sizes corresponds to their STROOPWAFEL weights. The inset colourbars indicate the percentage of the population represented by each contour and the annotated mass is the average chirp mass for all binaries in the panel, which is used in plotting the grid lines. We plot diagonal lines of constant distance, where the straight lines show the signal for a circular binary of average chirp mass, whilst the curved line shows the signal for an eccentric binary with $e = 0.98$. The vertical lines indicate the inspiral time for a circular binary with the average chirp mass. In the top panel we overlay the LISA verification binaries from \citet{Kupfer+2018}. In the bottom panels we add the 90\% contour line for only the circular sources to show how the distribution changes without eccentric sources present.}
    \label{fig:dcos_on_sc}
\end{figure*}

We show the expected distribution of detectable DCOs together with the LISA sensitivity curve in Figure~\ref{fig:dcos_on_sc}. This shows that the detectable population of these massive DCOs is concentrated at comparatively lower frequencies than the LISA verification binaries (shown as stars in the top panel). This is expected since producing the same SNR as a BHBH, BHNS or NSNS with a relatively lower mass (circular) WDWD requires a higher frequency. This finding is in agreement with \citet{Sesana+2020} and as noted in that work, this could possibly be used to distinguish these more massive DCOs from WDWDs probabilistically. There are several other notable features in these distributions and to understand these we overplot reference lines indicating where a circular binary with the average chirp mass (annotated in each panel) would reside for a given remaining inspiral time (vertical lines) and a given distance (diagonal lines).

We would expect that if a population was entirely circular, it should be bounded approximately between the $0.1$--$30 \unit{kpc}$ lines (roughly the minimum and maximum distance to a source in the Milky Way). From inspection of the bottom panels with each individual DCO type we see that, though this is the true for a large fraction of the population, there is a distinct subpopulation of binaries that extend downwards, especially around $2 \unit{mHz}$. This offshoot is composed of eccentric binaries for which the circular distance contours do not apply. For instance, we plot the 90\% contour of only the circular sources in our sample over the density distribution in each of the bottom panels and it is clear that the circular subsample is bounded by the distance lines. We also plot a line of constant distance at $30 \unit{kpc}$ for an eccentric binary with $e = 0.98$ to show the differences for eccentric sources. Additionally, we note that the peak of the density distribution coincides with the centre of the Milky Way as expected, since binaries are most likely to be formed towards the centre of the Galaxy.

@@

\begin{figure*}[p]
    \centering
    \includegraphics[width=\textwidth]{3_dco_detections.pdf}
    \caption{The number of expected detections in the LISA mission for different DCO types and model variations. Error bars show the 1- (solid) and 2-$\sigma$ (dotted) Poisson uncertainties. An arrow indicates that the error bar extends to zero. The left axis and grid lines show the number of detections in a 4-year LISA mission and the right axis shows an approximation of the number of detections in a 10-year mission (we scale the axis by $\sqrt{T_{\rm obs}}$, see Table~\ref{tab:detection_rates} for exact rates). Each model is described in further detail in Table~\ref{tab:physics_variations} and details of the fiducial assumptions are in Section~\ref{sec:fiducial_physics}. \todo{subject to change with the updated/new models and new uncertainty estimates}}
    \label{fig:detection_rates}
\end{figure*}
